% !TEX root = ../PhD Thesis.tex

\chapter*{Preface and Acknowledgments}
\addcontentsline{toc}{chapter}{Preface and Acknowledgments}

“Conformity is the death of progress.”
\\
\\
It was this motto that I wielded without ever yielding. I never truly felt like a scientist. I am disorganised, and I am especially sensitive to failure, all in a field where failure is not incidental but endemic. I have always resisted the canonical academic path, in which a scientist devotes an entire career to a single gene, protein, pathway, or at least a narrow thematic corridor. Instead, I always preferred ever-shifting terrain: exploratory projects, ever changing questions, intellectual migration.
\\
\\
I dislike repetition. I dislike the sensation of stagnant progress. The feeling that one sacrifices mental health for a marginal increment of data, like a World War I general sacrificing men for a few inches of land. The analogy is cold, disproportionate, and I recognise that fully. Yet the very impulse to make it is revealing.
\\
\\
Although I was a strong student overall, my highest marks were always in the humanities: languages and history. For many years, I wanted to be a creative writer, to craft novels with morally ambiguous characters, where the hero confronts a great external conflict only to discover that the true struggle is internal. How much of themselves is he willing to sacrifice for his goal? In that regard, life imitated art. My world, too, contains grey characters: stress granules as both tumour suppressors and tumour promoters. And though I am no hero, I have repeatedly faced a similar question. How much of myself should be sacrificed for scientific progress?
\\
\\
The project itself was not without turbulence. At its inception, it was challenged on conceptual grounds: the heterogeneity of stressors would inevitably produce heterogeneous stress granules. If the upstream perturbations vary widely, how could the downstream assemblies be consistent? My preliminary data, precisely the differential centrifugation studies, suggested otherwise. The granules appeared remarkably conserved.
\\
\\
Of course, there is a marked irony in this. The very data that secured the project’s legitimacy are now the data I regard as methodologically inadequate. That does not imply that stress granules induced by distinct stressors share nothing in common. It suggests only that the question is more complex than I once believed. I remain convinced that all knowledge, including “negative” knowledge, as was so called in regards to my project, as invaluable to scientific progress.
\\
\\
As a neurodivergent individual, I often felt out of place, though never entirely without one. I am obsessive. I cannot rest when a question lacks an answer. Many nights were spent carousing with insomnia. I am perhaps more imaginative than most, capable of generating hypotheses that bridge conceptual gaps others might not attempt to cross. Yet imagination is double-edged: when one sees connections everywhere, even the impossible can appear plausible. A hypothesis that felt visionary to me may have seemed quixotic to others.
\\
\\
I have always been competitive, but situated at a crossroads. I know less mathematics than engineers, less molecular biology than career biologists. Mediocre in many domains, but great at none. Watching peers excel in clearly defined strengths, I often felt diffuse by comparison. And so I leaned on other faculties: stubbornness, diligence, an unhealthy disregard for personal cost.
\\
\\
Within that darkness, I often took proverbial shots in the dark. I pursued techniques and ideas others would not entertain: simulating RNA sequencing for marginal informational gain; resurrecting equations older than myself to calculate how a single RNA transcript might sediment under specific centrifugation regimes, assuming, improbably, that it behaved as a sphere influenced only by mass; estimating how two molecular additions, individually negligible yet cumulatively significant, would alter that sedimentation. I explored how specific RNA-binding proteins regulate not only themselves but one another. Much of it bordered on the unreasonable. Most of it I did not particularly enjoy.
\\
\\
My preference has always been different: to use the keys I already possess to open new doors. Yet I was seldom afforded that luxury. Relentlessness became my default mode, less a virtue than a coping mechanism. Failure felt personal, with success merely feeling like a failure that was yet to come. Whenever a question was answered and closed, it simply brought relief, but never celebration.
\\
\\
Perhaps, when I read this again in the future, I will be more committed to the pursuit of doors rather than the compulsive forging of new keys. Or at least, I will believe that the keys I forged have contributed, however modestly, to the “bettering of humankind".
\\
\\
And now, an abrupt change, from the bleak paragraph (albeit less so than what is my personal signature), to the concomitant section of acknowledgments, which I chose to maintain herein.
\\
\\
First, I thank my thesis evaluating committee. I know that my writing is not always particularly enjoyable, especially in terms of length, and for your effort and dedication, I am truly grateful. 
\\
\\
Next, I must naturally thank my supervisor, Nuno, for taking a chance on someone with no formal programming experience whatsoever, and for giving me the freedom I craved. He allowed me to design my own research project and decide where I wanted to lay my tracks. I could always choose the direction, but with his help the train kept running smoothly. Not that the train never had problems. At one point it remained parked at an \textit{apeadeiro} for almost a year, as the integrity of the tracks could not be assured.
\\
\\
Next, I thank all the members of the Disease Transcriptomics Lab throughout the years, especially Mariana and Miguel, who helped me integrate when everything was still new. Jacek, for all the tequila we shared, and for all the pushups we did together; Diana, who so gently took over my project when my mental state crumbled completely, with whom I shared so many pet photos; Dani, for bringing a more nuanced philosophy outlook to the lab beyond my own; and Marta Bica, as my introverted companion in an extroverted lab.
\\
\\
I must also thank Vanessa and her lab for welcoming a computational biology PhD student into the wet lab after three years of rust. That was a bold move. I thank Vanessa in particular because I know I was not an easy student to supervise, but thanks to her guidance the work improved drastically. Without her persistence pushing me forward, I would probably never have uncovered the stealthy connection between stress granules and mitochondria.
I also thank the lab members: Elvira, for all the help with experiments and for the many conversations we shared about life and its many doubts; Ana, for helping me with experiments, especially during my induction and with the constant restocking of solutions and reagents; and both Ana and Bernardo for sharing so much time with me and indulging my strange hobbies. I leave here a promise: the legend of Tico and Teco will continue, and the “Gang da Capa Vermelha” will not prevail.
\\
\\
I would also like to thank the Lisbon BioMed PhD Program and all those responsible for its creation and maintenance. Many PhD programs treat candidates as tools, labour meant to execute predefined projects with which they have little connection. The Lisbon BioMed program did something unusual: it allowed us to enroll without a predefined PhD project. This allowed a natural match between candidates and projects, but also between candidates and supervisors. A name or a CV rarely captures the small details that make collaborations actually work, and it is a shame that this layer is so often ignored, especially on novel programs in the vicinity.
\\
\\
I also thank my colleagues from the program. Starting together meant struggling together. Suffering is easy, everyone can do it. Suffering alone is much harder. By sharing that journey we learned something important: most of the time, the struggles were not because of us. That realisation did wonders for starving our imposter syndrome instead of feeding it.
\\
\\
In this context, I must thank Bárbara and Marta, with whom I bonded while organising what I will confidently call the best CAML PhD meeting. Why we decided to take that responsibility upon ourselves remains unclear, but I am very glad we did. Despite the stress, the last-minute schedule changes, and the controlled chaos, the moments we shared are memories I will always cherish, along with the stickers, the mocktail recipes, and the legendary “1s.”
\\
\\
I thank Cláudia and Helena from the comms, for all the times they so gently provided a break, whenever rest was a necessity. The golf charity event is still etched in my memories.
\\
\\
From the Lisbon BioMed program I must also mention Eller, with whom we created a DnD group that played together many times. In science, our creative side is often malnourished, and having a space where imagination was not only allowed but encouraged was invaluable. The day we killed Santa will always remain one of my favorite memories.
\\
\\
I must also highlight the italian Marta for sharing the CAML organisation adventure, for the fun moments, and for the random dances during dinner that I still remember. And of course, for joining me in the questionable life choice of becoming a PhD representative. I also thank her for making me feel less alone in wanting to continue in academia, despite all available evidence suggesting that might not be the most rational decision.
\\
\\
There are also two people with whom I shared much of my life during these almost five years. Francisca, whose full name I will not write because it is as long as that of an old noble (which, coming from me, is admittedly ironic), for being a constant and loyal lion. We celebrated two championships together, many victories, and also our fair share of painful defeats. We even have that photo where nobody cared who we were before we put on the mask. Our scientific interests often overlapped as well: with her focus on cancer, we frequently collaborated as two different scientists pursuing the same goal.
\\
\\
And of course, I must mention Rita, with whom I started the PhD program at the same time, in the same lab. We laughed together many times and suffered together even more. The endless world of modules comes to mind, but to that I say "Bonne voyAGE!". I will always admire your sense of responsibility and the invisible work you do behind the scenes to keep the lab running smoothly. Our collaborations were especially fun, and I learned a lot about how a real programmer codes. After all, my workflow was always more end oriented. Why build complex functions when you can simply run the same function three times?
We also organised the CAML PhD meeting together and served as PhD representatives. It was not always easy. By nature I am competitive, and having someone who started so far ahead of me on the same path was sometimes discouraging. But in many ways it also helped, because by the time I reached those same points in the road, the path had already been cleared of pesky bureaucratic debris.
\\
\\
Moving outside of work-related friendships, I want to thank my long-time companions André, Francisco, and Rodrigo for remaining my friends through the years and for always welcoming me back exactly where we left off. The PhD forced many periods of intense work and long absences, but knowing I could return and pick things up as if no time had passed meant a lot to me. We played many games together, laughing both with and at each other. For me, those games were never just games, they were valuable moments of respite.
\\
\\
I must also mention my friends Pedro, the other Rita, and Gil, with whom I shared many moments: dinners, board games, terrible movies, New Year's Eve celebrations, and an epic DnD campaign. A chapter in that campaign lasted almost as long as my PhD, and choosing favorite moments is difficult. But the hunt for Derequi, and the revelations of Dorian and Llywelyn, remain particularly memorable. And of course, perhaps my happiest tradition, our halloween costume competition is one I hope to do for many more years.
\\
\\
Many other friends were also important to me, though over time I sadly drifted away from some of them. Just know that I will always remember you fondly, even if this is not the case for you.
\\
\\
Turning to my family, I must thank my uncles. It was thanks to them that I was able to move to Lisbon during my master's degree. They generously gave me a place to stay without asking for anything in return. Considering the current housing market, I would almost certainly not be here without them.
\\
\\
I also thank my mother for allowing me to come “so far away” in pursuit of better opportunities and my dreams, despite not being in medicine like she hoped so. I hope that being a doctor instead of a Dr. will suffice. 
\\
\\
And of course, I thank my sister, my eternal rival, and someone whose shadow I lived in for a long time. I was always destined to be number two: the second person in the family to obtain a master’s degree because you were the first, and hopefully the second to obtain a doctoral degree for the same reason. In the last year I chose a different publication name, but there will always have been a time when “A. Afonso" was simultaneously publishing about cancer and about the tristeza virus in oranges. I might not have gone into science if not for you. You always said that doing a PhD was a bad idea. In hindsight, you may have had a point.
\\
\\
I must also thank my newfound family, Vítor, Cristina, and Tomás, for always treating me as one of their own, constantly offering support whenever I needed. I am glad that now it is official.
\\
\\
My daughters and my son as well, for the constant purring and for always keeping me company. Lying on my lap, or on the keyboard at the most inconvenient possible moments, even when I worked well past any reasonable bedtime. Their furry companionship was always enough to warm me up, in every sense of the word.
\\
\\
Finally, and most importantly, I must thank my fiancée. I still hope that word remains “fiancée" and not “wife" by the time this is read. Otherwise it would mean I took far too long to finish. We were close before the PhD, and for a while the PhD pushed us apart. I am not always an easy person to love, and the PhD had a talent for amplifying my less charming qualities to their fullest expression. But we found our way back to each other, and that is not a small thing.
\\
\\
If anyone truly understands how hard these years were, it is you. Not because I was always good at showing it, but because you stayed close enough to see it regardless. You listened to every complaint, from sixteen-hour protocols to the metaphorical train stopping at the \textit{apeadeiro}, to the literal train stopping, to my deeply ingrained habit of placing work above everything else, including my own health and, too often, you. You never made me feel guilty for the long absences, the late nights, or the version of me that showed up exhausted and hollow and called it presence. You kept me grounded when I was at my most untethered, and sane to the most extent that I am capable of being sane. You also, and this matters more than it might seem, made me laugh, so so many times, which on the worst days was the only thing that worked.
\\
\\
I would not have finished this without you. I am not certain I would have wanted to. I owe you a debt that no dedication page and no acknowledgements section, however earnest, will ever come close to settling. But I intend to spend a long time trying.
\\
\\
And finally, for the two most important people, I leave a whole section for them.