% !TEX root = ../PhD Thesis.tex
\chapter{The WC Transcriptome}
\label{chap:ChapterIV}

In the \gls{sg} transcriptome chapter, we identified a technical batch effect introduced by the \gls{sg} isolation method. The \gls{wc} transcriptomes were not expected to carry such systematic biases, given the greater consistency of their collection procedure. However, multiple confounding variables could still influence the results, making the assessment of the \gls{wc} transcriptome less straightforward than initially anticipated.

\section{Materials and Methods}

\subsection{Dataset Procurement}

For dataset procurement, we queried the \gls{geo} Datasets repository using the term "stress granule", filtering for expression profiling by high-throughput sequencing and restricting results to human samples, following a strategy similar to that described in Section 4.1.1. As done for Chapter 5, we specifically targeted studies including both stressed and unstressed conditions, as well as experimental setups with and without \gls{sg} formation. 
\linebreak
\noindent
The following datasets, were used throughout this work:

\begin{itemize}
	\item GSE138988\cite{matheny_transcriptome-wide_2019};
	\item GSE119977\cite{matheny_rna_2021, matheny_corrigendum_2023};
	\item GSE173953\cite{paget_stress_2023}.
\end{itemize}
\noindent
All these datasets, or their joint integration, allowed for uncoupling of the transcriptomic effect of "stress" and/or genotype, from the effect of \gls{sg} assembly alone, which would otherwise confound the analyses.
\\
\linebreak
\noindent
During the literature review, we identified an unrelated dataset exhibiting a transcriptomic profile almost opposite to our putative \gls{sg} effect. To enable direct comparison between these two phenotypes, this senescent dataset (GSE236521\cite{lopez-polo_release_2024}) was also included.
\
\linebreak
\noindent
The batch correction methodology is described in the Results section of this chapter (Section 7.2.2).

\subsection{Dataset Pre-processing}

Downloaded raw reads (in FASTQ format) were trimmed using Trimmomatic (v0.39)\cite{bolger_trimmomatic_2014} operating in either \gls{se} or \gls{pe} mode, depending on the original study design, with appropriate adapter sequences and parameters. Trimmed reads were then aligned to the \gls{hg38} reference genome using STAR (v2.7.9a)\cite{dobin_star_2013} with the following parameters:

\begin{lstlisting}
	--outFilterType BySJout
	--alignSJoverhangMin 8
	--alignSJDBoverhangMin 1
	--alignIntronMin 20
	--alignIntronMax 1000000
	--alignMatesGapMax 1000000
	--outFilterMultimapNmax 1
	--outFilterMismatchNmax 999
	--outFilterMismatchNoverLmax 0.04
\end{lstlisting}
\noindent
Aligned reads (\gls{bam} files) were subsequently annotated using the GENCODE v42 primary assembly\cite{frankish_gencode_2021} and quantified with featureCounts (v2.0.3)\cite{liao_featurecounts_2014}.
\\
\linebreak
\noindent
For voyAGEr related pre-processing, please consult original publication\cite{schneider_voyager_2024}.

\subsection{Correlations}

Unless otherwise indicated, correlations shown in scatter plots and heatmaps were calculated using the t-statistic from differential expression, with Spearman's correlation coefficient $\rho$ and the associated test’s p-value. For correlations involving gene length, the canonical transcript length (in number of nucleotides), obtained from ENSEMBL\cite{cunningham_ensembl_2022} was used, with the longest transcript chosen in cases of ambiguity\cite{cunningham_ensembl_2022}. 

\subsection{DGEA}

To model expression changes between conditions, linear models were fitted to the gene expression data, incorporating genotype (\gls{eg} \gls{g3bp1} \gls{ko} or \gls{wt}), stress condition (\gls{eg} arsenite or sorbitol treatment), and \gls{sg} presence as covariates. Standard errors were subsequently moderated via empirical Bayes shrinkage\cite{smyth_linear_nodate}. Differentially expressed genes were defined separately for each dataset using dataset-specific log2FC values and B-statistics, rather than fixed thresholds.

\subsection{GSEA}

For each variable of interest, such as \gls{sg} effect, \gls{gsea}\cite{subramanian_gene_2005} was performed using the fgsea (v1.20.0)\cite{korotkevich_fast_2016} R package. Genes were ranked by t-statistics from differential expression analysis, ordered from most positive to most negative, and analysed against the Hallmark gene sets\cite{liberzon_molecular_2015}. The \gls{nes} was extracted to quantify the extent to which a given gene set is enriched at the extremes of the ranked gene list. A positive \gls{nes} indicates that the gene set is enriched amongst over-expressed genes, whereas a negative \gls{nes} indicates depletion.

\section{voyAGEr: a free web interface for the analysis of age-related gene expression alterations in human tissues}

To illustrate how even methodologically consistent comparisons, such as those relying on standard \gls{wc} \gls{rna} extraction procedures, can still yield substantial analytical challenges, a relevant side project warrants mention here. This involved the development of \textbf{voyAGEr}, an application to explore age-related gene expression alterations in \gls{gtex} samples, which resulted in a first-author publication\cite{schneider_voyager_2024}.



\subsection{Sample Origin as a Confounder}

The GTEx project, has two major sample types: Donor, and Post-Mortem (although this nonemclature is a bit problematic, as both sample types are effectively obtained post mortem). Donor samples are obtained from organ retrieval from organ donors following, for example, an accident. These tend to have a higher age variability, as both "young" and ''old`` individuals can have a mortal accident. Standard post mortem samples, are obtained from individuals who were already in an hospital setting, and eventually passed away. These are tipycally older. This means  that there is an oberrepresentation of donor samples at younger ages. Of course, if these samples are comparable, then, when properly accounting for other variables, one could still infer age-related gene expression alterations. Nevertheless, while in an hospital setting, samples are readily processed, when an accident occurs, personnel takes some time to arrive, and sample processing isnt as readily. This is obviously just an example of a difference, but still reveals substantial differences, across for example short transcripts, which are more prone to be affected by sample degradation. As such, even when comparing same aged samples from different origins, these genes will appear as differentially expressed. The problem is across the entire dataset, since both sample types arent balanced, this could give the impression that across ageing, shorter transcripts are preferentially expressed, difficulting any assessments related to age per se, which was correctly detected by reviewers during the submission process of the article.

\subsection{Batch Correction as a Solution}

Does this mean that GTEx dataset is useless to infer age related exoression changes? This would be a waste of such a rich dataset. After all, GTEx has many many samples, and with such  large number, one can assess what of the observed gene expresion alterations are given by the counfounding variables. Lets take the flagship example of this. Looking at the expression of the gene SFPQ, we can see two very clear modes, which correspond precisely to the two sample types we have. This is not surprising, as this gene is related to ventilator usage. Since of the sample types is associated presicely with this, we observe a massive difference between these. In the case of this gene, the difference is biologically real, and not owing to sample degradation or other non-biological technical issues. However, we can still consider that it is in fact a batch affect in this case, as it is only prsent in the batch of post-mortem samples. As such, it isnt wheter a variable is technical or biological in nature that inherently makes it a batch effect or not, but wheter it is a variable of interest for our biological question. If we are interested in seeing the effect of a treatment, and we have two different cell types, it isnt theoritically wrong to consider that in this case, the cell type is a batch effect, provided we dont care about cell type specific events, and only want what is in common in these cells. To solve these cases, and allow proper variable distribution of gene expression effects, one must therfore perform batch effect correction. To do this, the more samples, the merrier. Another issue, is that there must be no other variables confouned. If we have stressed neurons and unstressed fibroblasts, we cannot remove the "cell type" batch effect, as it is impossible to disantangle this effect from stress state in this case. Proper sample representation is necessary. In the case of a rich dataset like GTEx, this is asy to accomplish. Although several methods exist, we ended up developing our own for voyager. Simplyfying (for a more detailed explanation consult the original pubblication), we generate a linear model with our batch effect (sample type; donor vs postmortem) and our variables of interest (age and sex, in the case of voyager). From this linear model, we can thus obtain the gene expresion alterations that are explained by the batch effect, excluding those explained by our vriables of interest. We can then subtract these differences from our original expression matrix. Reapplying linear models with only our variables of interest in this modified gne expression matrix, will allow us to determine what gene expression changes are driven by our variables of interest. This ia a conservative approach, as we are only determining sample type gene expression changes. Another used method, such as combat, alllows to also determine sample type variation expression changes (ie, when the gane varies more in a sample type than in another). This is however, more violent, as we would also be potentially removing actual biological signal from our variables of interest. In voyager case, as our more conservativ approach seemingly resolved th isue, we prefered it, to maintain as much true biological signal as possible.

\section{What changes on a cell upon SG assembly?}

\subsection{SGs are anti-inflammatory}
As stated in the aims, as the majority of publicly available datasets do not purify SGs, infering their presence and effects in the cell from WC transcriptomes, would prove invaluable in studying their effect/function on other biological contexts. However, a major challenge in analyzing the WC transcriptome is separating the general stress response from SG-specific contributions. G3BP1/G3BP2 KO cells, which do not form visibly detectable SGs under oxidative stress, allowed us to distinguish these effects34,50,51. By comparing KO and WT cells under stressed and unstressed conditions, we isolated the contributions of SGs, stress, and the KO itself to the WC transcriptome (v. Methods)34,50,51. In this case, since techniques are theoritically the same, these datasets were all analysed together, after integration.In this case, since our sole variable of interest are transcriptomic changes driven by sg assembly, these other variables are, for our purpos, batch effects, and so, we used our previously method to correct for them. (in fact, this correction was not necessary, as the proper inclusion of these variables on the model, effectively controls for all these issues. Nevertheless, for the development of a SG signature, this process was easier to do if we had a gene expression matrix corrected for these other extraneous variables, and thus batch effect was effectly done here.) Contrary to their theorized function in fine-tuning the stress response, SGs appear to hamper this response, as evidenced by the downregulation of inflammation-related pathways (Supplementary Figure 4F), in a very consistent manner.  GSEA identified consistent downregulation of stress- and inflammation-related pathways in WC datasets, including interferon responses, TNF signaling via NF-κB, and IL6–JAK–STAT signaling (Figure 7B). Apoptosis pathways were also downregulated, supporting the proposed role of SGs as inhibitors of apoptosis. Functionally, SGs may act as “shock absorbers,” preventing excessive immune responses, a phenomenon previously observed by M. Paget with dsRNA7. Notably, similar patterns were observed regardless of the stressor, as the T. Matheny34,50,51 datasets used oxidative and hyperosmotic stress.
Looking directly at upregulated and downregulated genes, we observed downregulation of FOS, an immediate-early stress-responsive gene54, and upregulation of SERPINE1, a senescence-inducing gene and component of the senescence-associated secretory phenotype (SASP)14 (Figure 7C). 
Additionally, we observed consistent downregulation of several mitochondrially encoded genes (Figure 7C). Since one of their theorised functions is to sequester genes/proteins from pathways the "cell" doesnt want to be expressed at that moment, we considered that perhaps this WC transcriptome would exhibt a reverse correlation with our previously analysed SG transcriptome (although it isnt exactly clear whether or not SG sequestered RNAs are sequenced starting from regular WC rna extraction. It is plausible that they are removed on cleaning steps together with big organeless like mitochondria. Nevertheless, to our knowledge, this wasnt directly stated or tested, so its given here as an hypothesis. Nevertheless, as only a fraction of RNAs are in the SG, it is uncertain if their loss would lead to any substantial differences in the WC transcriptome.) Looking at these comparions, the PL dataset from SN. Somasekharan 202027 aligns more closely with WC transcriptomes than other SG datasets (Figure 7A), perhaps reflecting that only ~20\% of G3BP1 localizes to SGs. In contrast, the PL-derived dataset from Z. Ren 202352, targeting the same protein, does not, which challenges this interpretation. Nevertheless, as abovementioned, we do see a downregulation of MT rnas, which  had been previously observed to be enriched in several SGs, after proper correction.
To explore mechanisms underlying the dampening of stress and inflammatory signaling suggested by our meta-analysis, we compared our WC transcriptomes from SG-bearing cells to our SG-enriched transcripts. Althought these showed no overall correlation, mitochondrial RNAs were consistently anti-correlated (Figure 8A). 
\subsection{SGs have opposite roles to senescence.}

Looking at our results with SERPINE 1, and consistent with SG’s reported role as senescence inhibitors14, we found that a recent study targeting the SASP showed downregulation of pathways similar to those we observed in SG assembly63. However, a direct comparison reveals notable differences in gene expression between SASP inhibition and SG formation (Figure 8B).  Nevertheless, Both conditions shared downregulation of mitochondrially encoded RNAs, and although overall correlations were weak, GSEA revealed broad suppression of inflammation-related pathways (Figure 8C). Since both senescence and SGs are originated by stress exposure, it isnt too farfetched to say that they might be two different responses to the same stimuli, or that improper SG homeostasis can evolve into senescence.

